---
title: "Simulating study data: introduction"
author: "Keith S. Goldfeld"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulating study data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<STYLE TYPE="text/css">
<!--
  td{
    font-family: Arial; 
    font-size: 8pt;
    height: 2px;
    padding:0px;
    cellpadding="0";
    cellspacing="0";
    text-align: center;
  }
  th {
    font-family: Arial; 
    font-size: 9pt;
    height: 20px;
    font-weight: bold;
    text-align: center;
  }
  table { 
    border-spacing: 0px;
    border-collapse: collapse;
  }
--->
</STYLE>

```{r, echo = FALSE, message = FALSE}

library(simstudy)
library(ggplot2)
library(grid)
library(gridExtra)
library(knitr)

set.seed(33333)

opts_chunk$set(tidy.opts=list(width.cutoff=75), tidy=TRUE)

plotcolors <- c("#B84226", "#1B8445", "#1C5974")

cbbPalette <- c("#B84226","#B88F26", "#A5B435", "#1B8446",
                "#B87326","#B8A526", "#6CA723", "#1C5974") 

# 

ggtheme <- function(panelback = "white") {
  
  ggplot2::theme(
    panel.background = element_rect(fill = panelback),
    panel.grid = element_blank(),
    axis.ticks =  element_line(colour = "black"),
    panel.spacing =unit(0.25, "lines"),  # requires package grid
    panel.border = element_rect(fill = NA, colour="gray90"), 
    plot.title = element_text(size = 8,vjust=.5,hjust=0),
    axis.text = element_text(size=8),
    axis.title = element_text(size = 8)
  )  
  
}

splotfunc <- function(dt, ptitle) {

  dtplot <- dt[,.N,keyby=.(male, over65, rxGrp)][, .(rxGrp, grp = male * 2 + over65 * 1, N)]
  ggplot(dtplot, aes(factor(grp), N)) +
    geom_bar(aes(fill = factor(rxGrp)), alpha=.8, position = "dodge", stat="identity") +
    scale_fill_manual(values = plotcolors) +
    ggtitle(ptitle) +
    theme(legend.position = "none") +
    ggtheme() +
    xlab("Strata") +
    ylim(0,80)
}

aplotfunc <- function(dt, ptitle) {

  dtplot <- dt[,.N,keyby=.(rxGrp)]
  ggplot(dtplot, aes(factor(rxGrp), N)) +
    geom_bar(aes(fill = factor(rxGrp)), alpha=.8, position="dodge", stat="identity", width=.5) +
    scale_fill_manual(values = plotcolors) +
    ggtitle(ptitle) +
    theme(legend.position = "none") +
    ggtheme() +
    xlab("Treatment group") +
    ylim(0,150)
}

```

Simulation using `simstudy` has two primary steps. First, the user **defines** the data elements of a data set. Second, the user **generates** the data, using the definitions in the first step. Additional functionality exists to simulate observed or randomized **treatment assignment/exposures**, to generate **survival** data, to create **longitudinal/panel** data, to create **multi-level/hierarchical** data, to create datasets with **correlated variables** based on a specified covariance structure, to **merge** datasets, to create data sets with **missing** data, and to create non-linear relationships with underlying **spline** curves.

This vignette provides a brief introduction to the basics of generating data. For information on more elaborate data generating mechanisms, please visit  https://www.rdatagen.net/page/simstudy/.

## Defining the Data

The key to simulating data in `simstudy` is the creation of series of data definition tables that look like this:

```{r,  echo=FALSE}
def <- defData(varname = "nr", dist = "nonrandom", formula=7, id = "idnum")
def <- defData(def,varname="x1", dist="uniform", formula="10;20")
def <- defData(def,varname="y1", formula="nr + x1 * 2", variance=8)
def <- defData(def,varname="y2", dist="poisson", formula="nr - 0.2 * x1",link="log")
def <- defData(def, varname = "xnb", dist = "negBinomial" , formula="nr - 0.2 * x1", variance = 0.05, link = "log")
def <- defData(def,varname="xCat",formula = "0.3;0.2;0.5", dist="categorical")
def <- defData(def,varname="g1", dist="gamma", formula = "5+xCat", variance = 1, link = "log")
def <- defData(def,varname="b1", dist="beta", formula = "1+0.3*xCat", variance = 1, link = "logit")
def <- defData(def, varname = "a1", dist = "binary" , formula="-3 + xCat", link="logit")
def <- defData(def, varname = "a2", dist = "binomial" , formula="-3 + xCat", variance = 100, link="logit")

knitr::kable(def)
```

These *definition* tables can be generated two ways. One option is to to use any external editor that allows the creation of `csv` files, which can be read in with a call to `defRead`. An  alternative is to make repeated calls to the function `defData`. Here, we illustrate the R code that builds this definition table internally:

```{r, tidy = TRUE}
def <- defData(varname = "nr", dist = "nonrandom", formula=7, id = "idnum")
def <- defData(def,varname="x1",dist="uniform",formula="10;20")
def <- defData(def,varname="y1",formula="nr + x1 * 2",variance=8)
def <- defData(def,varname="y2",dist="poisson",formula="nr - 0.2 * x1",link="log")
def <- defData(def, varname = "xnb", dist = "negBinomial" , formula="nr - 0.2 * x1", variance = 0.05, link = "log")
def <- defData(def,varname="xCat",formula = "0.3;0.2;0.5",dist="categorical")
def <- defData(def,varname="g1", dist="gamma", formula = "5+xCat", variance = 1, link = "log")
def <- defData(def,varname="b1", dist="beta", formula = "1+0.3*xCat", variance = 1, link = "logit")
def <- defData(def, varname = "a1", dist = "binary" , formula="-3 + xCat", link="logit")
def <- defData(def, varname = "a2", dist = "binomial" , formula="-3 + xCat", variance = 100, link="logit") 
```

The first call to `defData` without specifying a definition name (in this example the definition name is *def*) creates a **new** data.table with a single row. An additional row is added to the table `def` each time the function `defData` is called. Each of these calls is the definition of a new field in the data set that will be generated. In this example, the first data field is named 'nr', defined as a constant with a value to be 7. In each call to `defData` the user defines a variable name, a distribution (the default is 'normal'), a mean formula (if applicable), a variance parameter (if applicable), and a link function for the mean (defaults to 'identity').\

The possible distributions include **normal**, **gamma**, **poisson**, **zero-truncated poisson**, **negative binomial**, **binary**, **binomial**, **beta**, **uniform**, **uniform integer**, **categorical**, and **deterministic/non-random**. For all of these distributions, key parameters defining the distribution are entered in the `formula`, `variance`, and `link` fields. 

In the case of the **normal**, **gamma**, **beta**, and **negative binomial** distributions, the formula specifies the mean. The formula can be a scalar value (number) or a string that represents a function of previously defined variables in the data set definition (or, as we will see later, in a previously generated data set). In the example, the mean of `y1`, a normally distributed value, is declared as a linear function of `nr` and `x1`, and the mean of `g1` is a function of the category defined by `xCat`. The `variance` field is defined only for normal, gamma, beta, and negative binomial random variables, and can only be defined as a scalar value. In the case of gamma, beta, and negative binomial variables, the value entered in variance field is really a dispersion value $d$. The variance of a gamma distributed variable will be $d \times mean^2$, for a beta distributed variable will be $mean \times (1- mean)/(1 + d)$, and for a negative binomial distributed variable, the variance will be $mean + d*mean^2$. \

In the case of the **poisson**, **zero-truncated poisson**, and **binary** distributions, the formula also specifies the mean. The variance is not a valid parameter in these cases, but the `link` field is. The default link is 'identity' but a 'log' link is available for the Poisson distributions and a "logit" link is available for the binary outcomes. In this example, `y2` is defined as Poisson random variable with a mean that is function of `nr` and `x1` on the log scale. For binary variables, which take a value of 0 or 1, the formula represents probability (with the 'identity' link) or log odds (with the 'logit' link) of the variable having a value of 1. In the example, `a1` has been defined as a binary random variable with a log odds that is a function of `xCat`.  \

In the case of the *binomial* distribution, the formula specifies the probability of success $p$, and the variance field is used to specify the number of trials $n$. The mean of this distribution is $n*p$, and the variance is $n*p*(1-p)$.
  
Variables defined with a **uniform**, **uniform integer**, **categorical**, or **deterministic/non-random** distribution are specified using the formula only. The `variance` and `link` fields are not used in these cases.  \
  
For a uniformly distributed variable, The formula is a string with the format "a;b", where *a* and *b* are scalars or functions of previously defined variables. The uniform distribution has two parameters - the minimum and the maximum. In this case, *a* represents the minimum and *b* represents the maximum.  \
  
For a categorical variable with \(k\) categories, the formula is a string of  probabilities that sum to 1: "\(p_1 ; p_2 ; ... ; p_k\)". \(p_1\) is the probability of the random variable falling category 1, \(p_2\) is the probability of category 2, etc. The probabilities can be specified as functions of other variables previously defined. In the example, `xCat` has three possibilities with probabilities 0.3, 0.2, and 0.5, respectively.  \
  
Non-random variables are defined by the formula. Since these variables are deterministic, variance is not relevant. They can be functions of previously defined variables or a scalar, as we see in the sample for variable defined as `nr`.

## Generating the Data

After the data set definitions have been created, a new data set with \(n\) observations can be created with a call to function **`genData`**. In this example, 1,000 observations are generated using the data set definitions in **`def`**, and then stored in the object **`dt`**:

```{r, tidy = TRUE}
dt <- genData(1000, def)
dt
```

New data can be added to an existing data set with a call to function **`addColumns`**. The new data definitions are created with a call to **`defData`** and then included as an argument in the call to **`addColumns`**:

```{r, tidy = TRUE}
addef <- defDataAdd(varname = "zExtra", dist = "normal", formula = '3 + y1', 
                 variance = 2)

dt <- addColumns(addef, dt)
dt
```

## Generating the Treatment/Exposure

Treatment assignment can be accomplished through the original data generation process, using `defData` and `genData`. However, the functions `trtAssign` and `trtObserve` provide more options to generate treatment assignment.

### Assigned treatment

Treatment assignment can simulate how treatment is made in a randomized study. Assignment to treatment groups can be (close to) balanced (as would occur in a block randomized trial); this balancing can be done without or without strata. Alternatively, the assignment can be left to chance without blocking; in this case, balance across treatment groups is not guaranteed, particularly with small sample sizes.

First, create the data definition:

```{r, tidy = TRUE}
def <- defData(varname = "male", dist = "binary", formula = .5 , id="cid")
def <- defData(def, varname = "over65", dist = "binary", formula = "-1.7 + .8*male", link="logit")
def <- defData(def, varname = "baseDBP", dist = "normal", formula = 70, variance = 40)

dtstudy <- genData(330, def)
```

*Balanced treatment assignment, stratified by gender and age category (not blood pressure)*
```{r, tidy = TRUE}
study1 <- trtAssign(dtstudy , n=3, balanced = TRUE, strata = c("male","over65"), grpName = "rxGrp")

study1
```

*Balanced treatment assignment (without stratification)*
```{r, tidy = TRUE}
study2 <- trtAssign(dtstudy , n=3, balanced = TRUE, grpName = "rxGrp")
```

*Random (unbalanced) treatment assignment*
```{r, tidy = TRUE}
study3 <- trtAssign(dtstudy , n=3, balanced = FALSE, grpName = "rxGrp")
```

*Comparison of three treatment assignment mechanisms*
```{r, tidy = TRUE, echo = FALSE, fig.width = 4, fig.height = 6}
p1 <- splotfunc(study1, "Balanced within strata")
p1a <- aplotfunc(study1, "")

p2 <- splotfunc(study2, "Balanced without strata")
p2a <- aplotfunc(study2, "")

p3 <- splotfunc(study3, "Random allocation")
p3a <- aplotfunc(study3, "")

grid.arrange(p1, p1a, p2, p2a, p3, p3a, ncol=2)
```

### Observed treatment

If exposure or treatment is observed (rather than randomly assigned), use `trtObserve` to generate groups. There may be any number of possible exposure or treatment groups, and the probability of exposure to a specific level can depend on covariates already in the data set. In this case, there are three exposure groups that vary by gender and age:

```{r, tidy = TRUE}
formula1 <- c("-2 + 2*male - .5*over65", "-1 + 2*male + .5*over65")
dtExp <- trtObserve(dtstudy, formulas = formula1, logit.link = TRUE, grpName = "exposure")
```

Here are the exposure distributions by gender and age:

```{r, tidy = TRUE, echo = FALSE, fig.width = 6.5, fig.height = 2.5}
dtplot1 <- dtExp[,.N,keyby=.(male,exposure)]
p1 <- ggplot(data = dtplot1, aes(x=factor(male), y=N)) +
  geom_bar(aes(fill=factor(exposure)), alpha = .8, stat="identity", position = "dodge") +
  ggtheme() +
  theme(axis.title.x = element_blank()) +
  theme(legend.title = element_text(size = 8)) +
  ylim(0, 150) +
  scale_fill_manual(values = plotcolors, name = "Exposure") +
  scale_x_discrete(breaks = c(0, 1), labels = c("Female", "Male")) +
  ylab("Number exposed")+
  ggtitle("Gender")

dtplot2 <- dtExp[,.N,keyby=.(over65,exposure)]
p2 <- ggplot(data = dtplot2, aes(x=factor(over65), y=N)) +
  geom_bar(aes(fill=factor(exposure)), alpha = .8, stat="identity", position = "dodge") +
  ggtheme() +
  theme(axis.title.x = element_blank()) +
  theme(legend.title = element_text(size = 8)) +
  ylim(0, 150) +
  scale_fill_manual(values = plotcolors, name = "Exposure") +
  scale_x_discrete(breaks = c(0, 1), labels = c("65 or younger", "Over 65")) +
  ylab("Number exposed") +
  ggtitle("Age")

grid.arrange(p1,p2,nrow=1)

```

Here is a second case of three exposures where the exposure is independent of any covariates. Note that specifying the formula as `c(.35, .45)` is the same as specifying it is `c(.35, .45, .20)`. Also, when referring to probabilities, the identity link is used:

```{r, tidy = TRUE}
formula2 <- c(.35, .45)

dtExp2 <- trtObserve(dtstudy, formulas = formula2, logit.link = FALSE, grpName = "exposure")
```

```{r, tidy = TRUE, echo = FALSE, fig.width = 6.5, fig.height = 2.5}
dtplot1a <- dtExp2[,.N,keyby=.(male,exposure)]
p1a <- ggplot(data = dtplot1a, aes(x=factor(male), y=N)) +
  geom_bar(aes(fill=factor(exposure)), alpha = .8, stat="identity", position = "dodge") +
  ggtheme() +
  theme(axis.title.x = element_blank()) +
  theme(legend.title = element_text(size = 8)) +
  ylim(0, 150) +
  scale_fill_manual(values = plotcolors, name = "Exposure") +
  scale_x_discrete(breaks = c(0, 1), labels = c("Female", "Male")) +
  ylab("Number exposed")+
  ggtitle("Gender")

dtplot2a <- dtExp2[,.N,keyby=.(over65,exposure)]
p2a <- ggplot(data = dtplot2a, aes(x=factor(over65), y=N)) +
  geom_bar(aes(fill=factor(exposure)), alpha = .8, stat="identity", position = "dodge") +
  ggtheme() +
  theme(axis.title.x = element_blank()) +
  theme(legend.title = element_text(size = 8)) +
  ylim(0, 150) +
  scale_fill_manual(values = plotcolors, name = "Exposure") +
  scale_x_discrete(breaks = c(0, 1), labels = c("65 or younger", "Over 65")) +
  ylab("Number exposed") +
  ggtitle("Age")

grid.arrange(p1a,p2a,nrow=1)
```
