---
title: "Survival Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Survival Data}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---



```{r, echo = FALSE, message = FALSE}
library(simstudy)
# library(ggplot2)
library(scales)
library(grid)
library(gridExtra)
library(survival)
library(gee)
library(data.table)
```

```{r, echo = FALSE}
plotcolors <- c("#B84226", "#1B8445", "#1C5974")

cbbPalette <- c("#B84226","#B88F26", "#A5B435", "#1B8446",
                "#B87326","#B8A526", "#6CA723", "#1C5974") 

ggtheme <- function(panelback = "white") {
  
  ggplot2::theme(
    panel.background = ggplot2::element_rect(fill = panelback),
    panel.grid = ggplot2::element_blank(),
    axis.ticks =  ggplot2::element_line(colour = "black"),
    panel.spacing = ggplot2::unit(0.25, "lines"),  # requires package grid
    panel.border = ggplot2::element_rect(fill = NA, colour="gray90"), 
    plot.title = ggplot2::element_text(size = 8,vjust=.5,hjust=0),
    axis.text = ggplot2::element_text(size=8),
    axis.title = ggplot2::element_text(size = 8)
  )  
  
}
```

Time-to-event data, including both survival and censoring times, are created using functions `defSurv` and `genSurv`. The survival data definitions require a variable name as well as a specification of a scale value, which determines the mean survival time at a baseline level of covariates (i.e. all covariates set to 0). The Weibull distribution is used to generate these survival times. In addition, covariates (which have been defined previously) that influence survival time can be included in the `formula` field. Positive coefficients are associated with longer survival times (and lower hazard rates). Finally, the *shape* of the distribution can be specified. A `shape` value of 1 reflects the *exponential* distribution. As of `simstudy` version 0.5.0, it is also possible to generate survival data that violate a proportional hazards assumption. In addition, data with two or more competing risks can be generated.

### Weibull distribution

The density, mean, and variance of the Weibull distribution that is used in the data generation process are defined by the parameters $\lambda$ (scale) and $\nu$ (shape) as shown below.

\begin{aligned}
f(t) &= \frac{t^{\frac{1}{\nu}-1}}{\lambda \nu} exp\left(-\frac{t^\frac{1}{\nu}}{\lambda}\right) \\
E(T) &= \lambda ^ \nu \Gamma(\nu + 1) \\
Var(T) &= (\lambda^2)^\nu \left( \Gamma(2 \nu + 1) - \Gamma^2(\nu + 1) \right) \\
\end{aligned}

<br>

The survival time $T$ data are generated based on this formula:

$$
T = \left( -\frac{log(U) \lambda}{exp(\beta ^ \prime x)}  \right)^\nu,
$$

where $U$ is a uniform random variable between 0 and 1, $\beta$ is a vector of parameters in a Cox proportional hazard model, and $x$ is a vector of covariates that impact survival time. $\lambda$ and $\nu$ can also vary by covariates.

### Generating standard survival data with censoring

Here is an example showing how to generate data with covariates. In this case the scale and shape parameters will vary by group membership.

```{r, tidy = TRUE}

# Baseline data definitions

def <- defData(varname = "x1", formula = .5, dist = "binary")
def <- defData(def,varname = "grp", formula = .5, dist = "binary")

# Survival data definitions

set.seed(282716)

sdef <- defSurv(varname = "survTime", formula = "1.5*x1", scale = "grp*50 + (1-grp)*25", shape = "grp*1 + (1-grp)*1.5")
sdef <- defSurv(sdef, varname = "censorTime", scale = 80, shape = 1)

sdef

```

The data are generated with calls to `genData` and `genSurv`:

```{r, tidy = TRUE}

# Baseline data definitions

dtSurv <- genData(300, def)
dtSurv <- genSurv(dtSurv, sdef)

head(dtSurv)

# A comparison of survival by group and x1

dtSurv[,round(mean(survTime),1), keyby = .(grp,x1)]

```

Observed survival times and censoring indicators can be generated using the competing risk functionality and specifying a censoring variable:

```{r, tidy = TRUE}
dtSurv <- genData(300, def)
dtSurv <- genSurv(dtSurv, sdef, timeName = "obsTime", 
            censorName = "censorTime", eventName = "status", 
            keepEvents = TRUE)

head(dtSurv)

# estimate proportion of censoring by x1 and group

dtSurv[,round(1-mean(status),2), keyby = .(grp,x1)]
```


Here is a Kaplan-Meier plot of the data by the four groups:


```{r, tidy = TRUE, echo = FALSE, fig.width = 6.5, fig.height = 3.5, warning=FALSE}
fit <- survfit(Surv(obsTime, status) ~ x1+grp, data=dtSurv)

survminer::ggsurvplot(fit, data = dtSurv,
  palette = cbbPalette,
  size = .5,
  ggtheme = ggtheme("grey94")
    # ggplot2::theme(axis.title = ggplot2::element_text(size = 9),
    #                        panel.grid = ggplot2::element_blank())
)
```

Here is a survival analysis (using a Cox proportional hazard model) of a slightly simplified data set with two baseline covariates only:

```{r, tidy = TRUE}

# Baseline data definitions

def <- defData(varname = "x1", formula = .5, dist = "binary")
def <- defData(def,varname = "x2", formula = .5, dist = "binary")

# Survival data definitions

sdef <- defSurv(varname = "survTime", formula = "1.5*x1 - .8*x2", scale = 50, shape = 1/2)
sdef <- defSurv(sdef, varname = "censorTime", scale = 80, shape = 1)

dtSurv <- genData(300, def)
dtSurv <- genSurv(dtSurv, sdef, timeName = "obsTime", 
            censorName = "censorTime", eventName = "status")

coxfit <- survival::coxph(Surv(obsTime, status) ~ x1 + x2, data = dtSurv)
```

The 95\% confidence intervals of the parameter estimates include the values used to generate the data:

```{r, echo=FALSE}
gtsummary::tbl_regression(coxfit)
```

### Competing risks

In the previous example, we actually used the competing risk mechanism in `genSurv` to generate an observed time variable (which was the earliest of the censoring and event time). This is done by specifying a *timeName* argument that will represent the observed time value. The event status is indicated in the field set by the *eventName* argument (which defaults to "event"). If a variable name is indicated in the *censorName* argument, the censored events automatically have a value of 0. As we saw above, competing risk information can be generated as part of `genSurv`. However, there is an additional function `addCompRisk` that will generate the competing risk information using an existing data set. The example here will take that approach.

```{r}
d1 <- defData(varname = "x1", formula = .5, dist = "binary")
d1 <- defData(d1, "x2", .5, dist = "binary")

dS <- defSurv(varname = "event_1", formula = "-10 - 0.6*x1 + 0.4*x2", shape = 0.3)
dS <- defSurv(dS, "event_2", "-6.5 + 0.3*x1 - 0.5*x2", shape = 0.5)
dS <- defSurv(dS, "censor", "-7", shape = 0.55)

dtSurv <- genData(1001, d1)
dtSurv <- genSurv(dtSurv, dS)

dtSurv
```

```{r}
dtSurv <- addCompRisk(dtSurv, events = c("event_1", "event_2", "censor"), 
            timeName = "time", censorName = "censor")
dtSurv
```

The competing risk data can be plotted using the cumulative incidence functions (rather than the survival curves):

```{r, tidy = TRUE, echo = FALSE, fig.width = 6.5, fig.height = 3.5, warning=FALSE}
fit <- survfit(Surv(time, event, type="mstate") ~ 1, data=dtSurv)
survminer::ggcompetingrisks(fit, ggtheme = ggtheme("grey94"))  + 
  ggplot2::scale_fill_manual(values = cbbPalette)
```

The data generation can all be done in two (instead of three) steps:

```{r}
dtSurv <- genData(101, d1)
dtSurv <- genSurv(dtSurv, dS, timeName = "time", censorName = "censor")
dtSurv
```

### Introducing non-proportional hazards



